---
title: "Flow Gauge Analysis"
author: "Lauren Puffer"
format: html
editor: visual
---

## Load packages

```{r}
library(here)
library(tidyverse)
library(janitor)
library(tmap)
library(terra)
library(tigris)
library(here)
library(sf)
library(stars)
```

## Load data

Data comes from 3 locations throughout the watershed.

```{r}
# load data from strawberry station
strawb_flow <- read.csv(here("data", "flow_gauges", "SF FEATHER R BL DIV DAM NR STRA.csv"))|>
  clean_names()

# load data from forbestown station
forbes_flow <- read.csv(here("data", "flow_gauges", "SF FEATHER R BL FORBESTOWN DAM.csv"))|>
  clean_names()

# load data from little grass valley station
grass_flow <- read.csv(here("data", "flow_gauges", "SF FEATHER R BL LITTLE GRASS VA.csv"))|>
  clean_names()
```

## Data cleaning

```{r}
# load lubridate
library(lubridate)

# Structure
# df$date <- mdy_hm(df$date)

# use lubridate to fix the dates in strawb_flow
strawb_flow$date <- mdy_hm(strawb_flow$date) # starts at 2001?

# use lubridate to fix the dates in forbes_flow
forbes_flow$date <- mdy_hm(forbes_flow$date)

# use lubridate to fix the dates in grass_flow
grass_flow$date <- mdy_hm(grass_flow$date)
```

Now we can get flows (discharge_m3_s) over time (date).

## Plot discharge over time

We will have to do this for each station. Then we will use patchwork to display them all at once. This will likely be a line graph!

```{r}
# load patchwork
library(patchwork)

# create line graph for strawb station
strawb_plot <- ggplot(strawb_flow, aes(x = date, y = discharge_m3_s)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "Flows near Strawberry Valley, CA",
    x = " ",
    y = expression(" ")
  ) +
  scale_x_datetime(date_labels = "%Y", date_breaks = "2 years")+
  theme_minimal()

# create line graph for forbes station
forbes_plot <- ggplot(forbes_flow, aes(x = date, y = dsicharge_m3_s)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "Flows near Forbestown, CA",
    x = " ",
    y = expression("Discharge (m"^3*"/s)")
  ) +
  scale_x_datetime(date_labels = "%Y", date_breaks = "2 years")+
  theme_minimal()

# create line graph for little grass valley station
grass_plot <- ggplot(grass_flow, aes(x = date, y = dishcarge_m3_s)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "Flows near Grass Valley, CA",
    x = "Year",
    y = expression(" ")
  ) +
    scale_x_datetime(date_labels = "%Y", date_breaks = "2 years")+
  theme_minimal()

# plot graphs together
discharge_plot <- strawb_plot/forbes_plot/grass_plot
```

Save the plot

```{r}
# save patchwork plot
ggsave(filename = here("plots", "discharge_by_station.png"),
  plot = discharge_plot,
  width = 10,
  height = 6,
  dpi = 300)
```

## Create a table of gauge stations

We will create our own dataset with information about the flow gauges.

```{r}
# create a data frame manually with site name, site id, and period of record
flow_station_df <- data.frame(
  site_id = c("11395200", "11396200", "11395030"),
  site_name = c("South Fork Feather River Below Diversion Dam Near Strawberry Valley CA ",
                "South Fork Feather River Below Forbestown Dam", "South Fork Feather River Below Little Grass Valley"),
  period_of_record = c("1961–present", "1961–present", "1961–present"),
  stringsAsFactors = FALSE
)

# View the dataframe
print(flow_station_df)
```

Create a table of flow gauge stations and period of record.

```{r}
# Load required packages
library(knitr)
library(kableExtra)

# Create a styled table
flow_station_df %>%
  kable(
    col.names = c("Site ID", "Site Name", "Period of Record"),
    caption = "Flow Gauge Stations"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  row_spec(0, extra_css = "font-family: 'EB Garamond', serif;") %>%
  row_spec(1:3, extra_css = "font-family: 'EB Garamond', serif;")

```

Now, we want to get the means, mins and max's of our flow station data by station.

```{r}
# make sure discharge is numeric
strawb_flow |>
  mutate(discharge_m3_s = as.numeric(discharge_m3_s))

forbes_flow <- forbes_flow |>
  mutate(discharge_m3_s = as.numeric(dsicharge_m3_s))

grass_flow <- grass_flow |>
  mutate(discharge_m3_s = as.numeric(dishcarge_m3_s))

# get flow means
mean_strawb <- mean(strawb_flow$discharge_m3_s)
mean_forbes <- mean(forbes_flow$discharge_m3_s)
mean_grass <- mean(grass_flow$discharge_m3_s)

# get the mins
min_strawb <- min(strawb_flow$discharge_m3_s)
min_forbes <- min(forbes_flow$discharge_m3_s)
min_grass <- min(grass_flow$discharge_m3_s)


# get the max
max_strawb <- max(strawb_flow$discharge_m3_s)
max_forbes <- max(forbes_flow$discharge_m3_s)
max_grass <- max(grass_flow$discharge_m3_s)
```

Make a summary dataframe.

```{r}
# make a summary of all mean, min, and max
# Combine your summary statistics into a data frame
flow_summary <- data.frame(
  site_name = c("South Fork Feather River Below Diversion Dam Near Strawberry Valley CA ",
                "South Fork Feather River Below Forbestown Dam", "South Fork Feather River Below Little Grass Valley"),
  mean_flow_m3_s = c(mean_strawb, mean_forbes, mean_grass),
  min_flow_m3_s = c(min_strawb, min_forbes, min_grass),
  max_flow_m3_s = c(max_strawb, max_forbes, max_grass)
)
```

Now we can make the table.

```{r}
# Create a nicely formatted table
flow_summary %>%
  kable(
    col.names = c("Site Name", "Mean Flow (m³/s)", "Min Flow (m³/s)", "Max Flow (m³/s)"),
    caption = "Summary of Flow Statistics by Site"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  row_spec(0, extra_css = "font-family: 'EB Garamond', serif;") %>%
  row_spec(1:nrow(flow_summary), extra_css = "font-family: 'EB Garamond', serif;")

```

## Flow gauge location

```{r}
# add latitude and longitude to flow_station_df
flow_station_df$latitude <- c(39.6475,39.5514, 39.7239 )

flow_station_df$longitude <- c(-121.1178, -121.2083,-121.0211 )
```

Change these points to geometries.

```{r}
# make geometries from lat and longs
df_sf <- st_as_sf(flow_station_df, coords = c("longitude", "latitude"), crs = 4267) # NAD 27 used in USGS flow station data

# get base map of california
library(tigris)

# get 'california' from tigris
# Load California county boundaries with the state's shapefile
ca_counties <- counties(cb = TRUE, state = "CA") |>
  st_transform(crs = 4267) # project to NAD 27

# filter for counties in watershed
watershed_counties <- ca_counties |>
  filter(NAME %in% c("Butte", "Plumas", "Sierra", "Lassen", "Yuba"))

# plot counties and flow stations
library(tmap)

flow_gauge_map <- tm_shape(watershed_counties)+
  tm_polygons() +
  tm_shape(df_sf)+
  tm_dots(fill = "red")
```

Get map of HUC 8.

```{r}
# bring in shape file of HUC 18020123
hucs <- st_read(here("watershed_shape","cat.shp"))

# check crs of points
st_crs(df_sf) # 4267

# check crs of watershed shape file
st_crs(hucs) # NA

# hucs are in 4269
st_crs(hucs) <- 6318

# check that they match
st_crs(df_sf_proj) == st_crs(hucs)

# transform to match the crs of counties
hucs_proj <- st_transform(hucs, st_crs(watershed_counties))
df_sf_proj <- st_transform(df_sf, st_crs(watershed_counties))



# filter for our HUC
huc_18020123 <- hucs_proj |>
  filter(HUC8 == "18020123")
```

Get a map of the watershed and flow gauge stations.

```{r}
# make mape of flow stations
flow_map <- tm_shape(huc_18020123) +
  tm_polygons(fill = "lightblue") +
  tm_shape(df_sf_proj) +
  tm_dots(col = "red", size = 0.2)

flow_map
```

Export csv

```{r}
# write csv
write.csv(flow_station_df, here("data", "flow_stations.csv"))
```

# Point sources

Load data

```{r}
# read in point source data
pt_source <- read.csv(here("data", "PCS_data_MF.csv")) |>
  clean_names()

# make geometries out of data frame
points_sf <- st_as_sf(pt_source,
                      coords = c("facility_longitude", "facility_latitude"),
                      crs = 3857) # 3857 is the EPSG code for Web Mercator lat/long

# bring in shape file of watershed
hucs <- st_read(here("watershed_shape","cat.shp"))
st_crs(hucs) <- 26910 # UTM Zone 10


# filter for our HUC
huc_18020123 <- hucs |>
  filter(HUC8 == "18020123")

# transform to fit points
points_proj <- st_transform(points_sf, 26910)

# check that CRSs match
st_crs(points_proj) == st_crs(huc_18020123)
```

## Troubleshooting

```{r}
# check bbox
st_bbox(huc_18020123)
st_bbox(points_proj)

# see 
```

## Plot point sources

Now we can attempt to plot these two on top of eachother (watershed boundary and point sources).

```{r}
tmap_mode("plot")


tm_shape(huc_18020123) +
  tm_polygons() +
  tm_shape(points_proj) +
  tm_dots(fill = "red", size = 0.5)

# try plotting just the points
plot(st_geometry(points_proj), add = TRUE, col = "red", pch = 20)

```
